He-Man: A domain specific language for lightweight concurrency

### 2012-03-11 paper/paper.tex
Finished introduction.

Our contribution is to distill the Lauer-Needham duality to its essence, by
compiling from a simple thread abstraction to self-contained event-driven C
code. Our language, He-Man: Haskell Event Manager, Apropos Networking, is a
compiled domain specific language embedded in Haskell, which offers a small
monadic language supporting thread spawning, blocking events, and external
function calls.

%XXX

The He-Man system consists of a compiler written in Haskell, and a lightweight
runtime in C, which links with the compiler's output. Both are small and easily
extensible. We target C because it is commonly used for network applications,

so it is trivial to link our

any code a programmer might prefer to write directly in C. 
Our vision is that C programmers can use He-Man for their
core application logic, and can write the rest of their application in C.

He-Man programs are compact and measurements indicate that it performs
acceptably.

\section{The event loop model}
Network applications require a very constrained form of
concurrency. Because in most network applications, I/O (both network
and disk) latency dominates the computation time, each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

In the event loop model, network I/O is done using non-blocking
\t{read} and \t{write} calls and all of the blocking for I/O is
centralized in a single call to a multiplexing system call such as
\t{select} or \t{epoll}. The multiplexing call will block until there
is some I/O ready to be performed and indicate to the application
which file descriptors are ready. From this, the application then
needs to decide how handle these events. This requires, for each
connection, having some sort of state machine to track what state the
connection is in and what needs to happen next, as well as other local
state.

@@

The event loop model is then, essentially, built around closure
converted trampolining continuation passing style. These are well
studied problems and can easily be done mechanically in the He-Man
backend.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{128kb_graph.png}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{16kb_graph.png}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}

>>>>

Our contribution is to distill the Lauer-Needham duality to its essence, by
compiling from a simple thread abstraction to self-contained event-driven C
code. This abstraction, He-Man: Haskell Event Manager, Apropos Networking, is a
compiled domain specific language embedded in Haskell, which offers a small
monadic language supporting thread spawning, blocking events, and external
function calls.

The He-Man system consists of a compiler written in Haskell, and a lightweight
runtime in C which links with the compiled code. He-Man allows programmers to
develop scalable network applications by expressing their I/O logic in He-Man,
and linking the resulting event loop to the rest of their application, which can
be written in C or any language which interfaces with C. We target C because it
is commonly used in this domain, and well-suited to performance-critical
applications.

In the remainder of the paper, we discuss the event-driven programming model,
our implementation of the He-Man system, and verify that its performance is
competitive with handwritten event loops.

\section{The event-driven model}

Network applications require a very constrained form of
concurrency. In most network applications, I/O (both network
and disk) latency dominates computation time, so each client's thread
of execution spends most of its time waiting on I/O. In particular, we
would like to always yield to a client no longer blocked on I/O; we
need not yield to blocked clients, nor do we (usually) need arbitrary
interleaving of execution.

In the event loop model, network I/O is done using non-blocking
\t{read} and \t{write} calls and all of the blocking for I/O is
centralized in a single call to a multiplexing system call such as
\t{select} or \t{epoll}. The multiplexing call will block until some 
I/O is ready to be performed, and indicates to the application
which file descriptors are ready. To decide how to handle these 
events, an application must, for each connection, have some sort of 
state machine to track what state the connection is in and what 
needs to happen next, as well as other local state.

@@

The event loop model is then, essentially, built around closure-
converted trampolining continuation passing style. These are well
studied problems and can easily be done mechanically in the He-Man
backend.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_100.pdf}
\caption{Web server benchmark, 100 B file.}
\label{fig:graph100}
\end{figure*}
